Notes and code examples to accompany "If the Null Fits, You Must Omit

### 2025-12-09 GroupPolarizationStatmod/code_examples.qmd
working on stats inference

---
title: Notes and code examples to accompany "If the Null Fits, You Must Omit"
author: Matt Turner · [https://mat.phd](https://mat.phd)
date: today
toc: true
editor: 
  markdown: 
    wrap: 90
bibliography: this.bib
---

@@

![](images/clipboard-1443885641.png){width="628"}

We call this a *null polarization model* (defined in terms of latent opinions) that
generates *spurious group polarization.* *Spurious group polarization* is when the
observed opinion distributions appear to polarize (i.e., *radicalize*, or *become more
extreme*) that does not represent the true change in latent opinions. The measurement
procedure *clips* extreme opinions less than the minimum bin value, or greater than the
maximum, causing the average observed opinion to *spuriously* appear polarized.

@@

```{r warning=FALSE}
# This is how I set the working directory
# Adapt for your own computer
setwd("~/workspace/gp-statmod/GroupPolarizationStatmod/")
# Load makeProbVec from file model.R
source("model.R")
```

@@

>>>>

---
title: Notes and code examples to accompany "If the Null Fits, You Must Omit"
author: Matt Turner · [https://mat.phd](https://mat.phd)
date: today
toc: true
editor: 
  markdown: 
    wrap: 90
bibliography: "paper/this.bib"

execute:
  # echo: false     # hide code
  warning: false  # hide warnings
  message: false  # hide messages
  # error: false
---

@@

![**Mean group opinions pre- and post-discussion from Moscovici & Zavalloni
(1969).**](images/clipboard-1443885641.png){width="575"}

We call this a *null polarization model* (defined in terms of latent opinions) that
generates *spurious group polarization.* *Spurious group polarization* is when the
observed opinion distributions appear to polarize (i.e., *radicalize*, or *become more
extreme*) that does notrepresent the true change in latent opinions. The measurement
procedure *clips* extreme opinions less than the minimum bin value, or greater than the
maximum, causing the average observed opinion to *spuriously* appear polarized.

@@

```{r warning=FALSE}
# This is how I set the working directory
# Adapt for your own computer
# setwd("~/workspace/gp-statmod/GroupPolarizationStatmod/")
# Load makeProbVec from file model.R
source("model.R")
```

@@

# The Generative Model and Bayesian Inference for False Detections

The analysis of the generative model has two major components:

1.  **Calculate false positive/detection rates for the simulated null polarization
    model**, *mere agreement*—where polarization experiment participant opinions agree
    more following discussion than before. Recall this is modeled as a decrease in the
    standard deviation of latent opinions, but a constant mean latent opinion.
2.  **Find an effect–detection threshold**, $d^*(\alpha)$, such that, under the null
    model, the probability of observing an estimated effect size exceeding
    \$d\^\*(\\alpha)\$ is \$\alpha\$: $$
     \Pr\!\left( \hat{d} \ge
     d^{*}(\alpha) \mid \mathcal{M}_{0} \right) = \alpha.
     $$ 
     This yields a null-model–calibrated false–positive rate for detecting a group
    polarization effect.

## Simulate each experimental design's false positive rate

Here's how to calculate the Type I or family-wise error rate for all the experiments
(though I misnamed the function back in the day, so it calls all of the error rates
"family-wise" because I had understood them to be synonyms).

```{r}
source("src/fwer.R")

probit_fits_dir <- "data/probit_fits"
# Set a "moderate" significance, according to Cohen's definitions 
significance_val <- 0.5
fwer_tbl <- calculate_fwer(probit_fits_dir, significance_val);

print(head(fwer_tbl))
```

This is used inside the `calc_fdr_vs_significance` function that adds another column to
this dataframe, the false discovery rate (FDR) for a set of significance values. Here it
is being called with the default metascience parameters with a base rate of $b=0.1$ and a
statistical power of $W=0.8$. We consider three effect sizes in this paper according to
the : "small", "medium", and "large" effect sizes traditionally assumed practically
meaningful across social sciences [@CohenBook1988]. I put this function in the `scripts` directory because to run the full analyses presented in the paper it's only necessary to run `calc_fdr_vs_significance` since it creates a column for the false positive rate and the estimated false detection rate calculated for each experimental design we studied.

This is reported for a variety of effect-detection thresholds, $\delta^*$, which the function name implies are "significance" values. When I first wrote this function
which are really effect-detection thresholds—so the code slightly misuses the term "significance", which typically refers only to a p-value based effect-detection threshold in the *null-hypothesis significance testing* approach. I updated the prose in the article to reflect this distinction, but left the code as-is. I again call the function with the defaults for illustration.

```{r}
source("scripts/analysis.R")

fdr_over_thresholds <- calc_fdr_vs_significance(
  probit_fits_dir, 
)
```

## Find effect-detection thresholds to achieve $\alpha = 0.05$

In the second step of our generative model analysis I calculate
